|

### 2024-05-03 manuscript/paper.Rmd
small updated.

\newpage

@@

`r gjm('You don't say "He says...." in a paper".  You say: "[citation] says..."'  )`

@@

@@

# Descriptive statistics

After removing 4 players who were missing heights, there were `r nrow(dat)` players included in our analysis.  Of these players, on average, enforcers in this data set tended to be older than the non-enforcers (EF: `r summ$mnage[summ$EF]` vs non-EF: `r summ$mnage[!summ$EF]` and EP: `r summEP$mnage[summEP$EP]` vs non-EP: `r summEP$mnage[!summEP$EP]`).  Additionally, as seen in Tables \ref{EDAEF} and \ref{EDAEP} enforcers tended to also be taller and heavier on average than non-enforcers

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EF = TRUE & EF = FALSE & p-value \\
 \hline
 Age (years) & `r summ$mnage[summ$EF]` (`r summ$seage[summ$EF]`) & `r summ$mnage[!summ$EF]` (`r summ$seage[!summ$EF]`) & `r ifelse(round(t.test(dat$age~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
 Weight (lbs) & `r summ$mnwt[summ$EF]` (`r summ$sewt[summ$EF]`) & `r summ$mnwt[!summ$EF]` (`r summ$sewt[!summ$EF]`) & `r ifelse(round(t.test(dat$Wt~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
  Height (in) & `r summ$mnht[summ$EF]` (`r summ$seht[summ$EF]`) & `r summ$mnht[!summ$EF]` (`r summ$seht[!summ$EF]`) & `r ifelse(round(t.test(dat$Ht~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EF)$p.value,2))`\\
    Goals  & `r summ$mngoals[summ$EF]` (`r summ$segoals[summ$EF]`) & `r summ$mngoals[!summ$EF]` (`r summ$segoals[!summ$EF]`) & `r ifelse(round(t.test(dat$G~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EF)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}
\label{EDAEF}
\caption{test}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EP = TRUE & EP = FALSE & p-value \\
 \hline
 Age (years) & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Weight (lbs) & `r summEP$mnwt[summEP$EP]` (`r summEP$sewt[summEP$EP]`) & `r summEP$mnwt[!summEP$EP]` (`r summEP$sewt[!summEP$EP]`) & `r ifelse(round(t.test(dat$Wt~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
  Height (in) & `r summEP$mnht[summEP$EP]` (`r summEP$seht[summEP$EP]`) & `r summEP$mnht[!summEP$EP]` (`r summEP$seht[!summEP$EP]`) & `r ifelse(round(t.test(dat$Ht~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EP)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}
\label{EDAEP}
\caption{Here is the caption for this table}
\end{table}

@@

```{r fig:KMEF, echo = FALSE, fig.cap="Kaplan Meier plot for EF" }
library(ggsurvfit)
survfit2(Surv ~ EFchar, data = dat) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) +
  add_confidence_interval()
```

```{r fig:KMEp, echo = FALSE, fig.cap="Kaplan Meier plot for EP" }
library(ggsurvfit)
survfit2(Surv ~ EPchar, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

Figures \@ref(KMEF) and \@ref(KMEP) show Kaplan-Meier plots for EF and EP

# Methods {#sec:methods}
In the matched cohort study presented in @Popkin2023, enforcers were matched to non-enforcers on birth year, weight, height, etc.  After matching, mortality rates between the two groups were computed.  Neither definition of enforcers (i.e. E-F or E-P) yielded significant difference in the mortality rates between enforcers and non-enforcers.

<!-- (Greg comment: After matching you CAN compare mortality rates and that comparison would be valid AT THAT GIVEN POINT when they are compared.  But a better analysis would be to use survival analysis.  The problem with mortality rate is that eventually everyone in both groups will be dead if you wait long enough and that comparison is meaningless. Survival times are much better comparison. ) -->

However, the authors of @Popkin2023 next conditioned on death yielding a sample size of n = 13 and n = 14 for the enforcer (defined by total fights) vs non-enforcer group, respectively, and show that the age at death is about 10 years less for the enforcers group relative to the non-enforcer group (p = 0.03).  This approach is not statistically valid, since once death is conditioned upon, the matching that is used to balance the two groups is broken and they can become imbalanced, which is what we believe the authors in @Popkin2023 observed.

This current manuscript aims to perform a re-analysis of the original study using survival analysis techniques, specifically Cox proportional hazards models, to look for differences in survival times of enforcers versus non-enforcers.

@@

# Results {#sec:results}
Two Cox proportional hazards models [CITATION] were fit: one for each definition of enforcer defined in @Popkin2023 (i.e. E-F and E-P).  In addition, to indicators for each definition of enforcer, games played (GP), height (Ht), weight (Wt), year of birth (DOB), and position (Center (C), Wing (W), or Defender(D)) were added to each model as control variables.

@@

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon Feb 26 09:06:03 2024 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & $\beta$ & $e^{\beta}$ & s.e. & p-value \\ 
  \hline
E-F TRUE & 0.17328 & 1.18920 & 0.28884 & 0.54856 \\ 
  GP & -0.00059 & 0.99941 & 0.00022 & 0.00815 \\ 
  Ht & -0.04484 & 0.95615 & 0.04430 & 0.31144 \\ 
  Wt & 0.02274 & 1.02300 & 0.00652 & 0.00049 \\ 
  DOB & -0.01766 & 0.98250 & 0.00821 & 0.03140 \\ 
  $I_D$ & -0.36479 & 0.69434 & 0.17428 & 0.03633 \\ 
  $I_W$ & -0.25905 & 0.77178 & 0.15869 & 0.10258 \\ 
   \hline
\end{tabular}
\caption{E-F}
\label{tab:EF}
\end{table}

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon Feb 26 09:04:35 2024 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & $\beta$ & $e^{\beta}$ & s.e. & p-value \\ 
  \hline
E-P TRUE & 0.83023 & 2.29385 & 0.26443 & 0.00169 \\ 
  GP & -0.00051 & 0.99949 & 0.00021 & 0.01759 \\ 
  Ht & -0.03859 & 0.96215 & 0.04406 & 0.38111 \\ 
  Wt & 0.02001 & 1.02022 & 0.00654 & 0.00221 \\ 
  DOB & -0.01778 & 0.98238 & 0.00827 & 0.03150 \\ 
  $I_D$ & -0.36900 & 0.69142 & 0.17444 & 0.03440 \\ 
  $I_W$ & -0.28804 & 0.74973 & 0.15946 & 0.07087 \\ 
   \hline
   \hline
\end{tabular}
\caption{E-P}
\label{tab:EP}
\end{table}

>>>>

@@

`r gjm('You dont say "He says...." in a paper".  You say: "[citation] says..."'  )`

@@

<!-- In the matched cohort study presented in @Popkin2023, enforcers were matched to non-enforcers on birth year, weight, height, etc.  After matching, mortality rates between the two groups were computed.  Neither definition of enforcers (i.e. E-F or E-P) yielded significant difference in the mortality rates between enforcers and non-enforcers.   -->

<!-- <!-- (Greg comment: After matching you CAN compare mortality rates and that comparison would be valid AT THAT GIVEN POINT when they are compared.  But a better analysis would be to use survival analysis.  The problem with mortality rate is that eventually everyone in both groups will be dead if you wait long enough and that comparison is meaningless. Survival times are much better comparison. ) --> -->

<!-- However, the authors of @Popkin2023 next conditioned on death yielding a sample size of n = 13 and n = 14 for the enforcer (defined by total fights) vs non-enforcer group, respectively, and show that the age at death is about 10 years less for the enforcers group relative to the non-enforcer group (p = 0.03).  This approach is not statistically valid, since once death is conditioned upon, the matching that is used to balance the two groups is broken and they can become imbalanced, which is what we believe the authors in @Popkin2023 observed.  -->

@@

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EF = TRUE & EF = FALSE & p-value \\
 \hline
 n & `r summ$n[summ$EF]` & `r summ$n[!summ$EF]` & \\
 Age (years) & `r summ$mnage[summ$EF]` (`r summ$seage[summ$EF]`) & `r summ$mnage[!summ$EF]` (`r summ$seage[!summ$EF]`) & `r ifelse(round(t.test(dat$age~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
 Weight (lbs) & `r summ$mnwt[summ$EF]` (`r summ$sewt[summ$EF]`) & `r summ$mnwt[!summ$EF]` (`r summ$sewt[!summ$EF]`) & `r ifelse(round(t.test(dat$Wt~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EF)$p.value,2))`\\
  Height (in) & `r summ$mnht[summ$EF]` (`r summ$seht[summ$EF]`) & `r summ$mnht[!summ$EF]` (`r summ$seht[!summ$EF]`) & `r ifelse(round(t.test(dat$Ht~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EF)$p.value,2))`\\
    Goals  & `r summ$mngoals[summ$EF]` (`r summ$segoals[summ$EF]`) & `r summ$mngoals[!summ$EF]` (`r summ$segoals[!summ$EF]`) & `r ifelse(round(t.test(dat$G~dat$EF)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EF)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}
\caption{test \label{tab:EDAEF}}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 Variable & EP = TRUE & EP = FALSE & p-value \\
 \hline
 n & `r summEP$n[summEP$EP]` & `r summEP$n[!summEP$EP]` & \\
  GP & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Age (years) & `r summEP$mnage[summEP$EP]` (`r summEP$seage[summEP$EP]`) & `r summEP$mnage[!summEP$EP]` (`r summEP$seage[!summEP$EP]`) & `r ifelse(round(t.test(dat$age~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
 Weight (lbs) & `r summEP$mnwt[summEP$EP]` (`r summEP$sewt[summEP$EP]`) & `r summEP$mnwt[!summEP$EP]` (`r summEP$sewt[!summEP$EP]`) & `r ifelse(round(t.test(dat$Wt~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$age~dat$EP)$p.value,2))`\\
  Height (in) & `r summEP$mnht[summEP$EP]` (`r summEP$seht[summEP$EP]`) & `r summEP$mnht[!summEP$EP]` (`r summEP$seht[!summEP$EP]`) & `r ifelse(round(t.test(dat$Ht~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$Ht~dat$EP)$p.value,2))`\\
      Goals  & `r summEP$mngoals[summEP$EP]` (`r summEP$segoals[summEP$EP]`) & `r summEP$mngoals[!summEP$EP]` (`r summEP$segoals[!summEP$EP]`) & `r ifelse(round(t.test(dat$G~dat$EP)$p.value,2) <0.0001,"<0.0001",round(t.test(dat$G~dat$EP)$p.value,2))`\\
 \hline
\end{tabular}
\end{center}

\caption{Here is the caption for this table \label{tab:EDAEP}}
\end{table}

After removing 4 players who were missing heights, there were `r nrow(dat)` players included in our analysis.  Of these players, on average, enforcers in this data set tended to be older than the non-enforcers (EF: `r summ$mnage[summ$EF]` vs non-EF: `r summ$mnage[!summ$EF]` and EP: `r summEP$mnage[summEP$EP]` vs non-EP: `r summEP$mnage[!summEP$EP]`).  Additionally, as seen in Tables \ref{tab:EDAEF} and \ref{tab:EDAEP} enforcers tended to also be taller and heavier on average than non-enforcers. The means of age, weight, height and goals are all significantly different between the enforcers and non-enforcer groups for both defintions of enforcers.

@@

```{r KMEF, echo = FALSE, fig.cap="Kaplan Meier plot for EF" }
library(ggsurvfit)
survfit2(Surv ~ EFchar, data = dat) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) +
  add_confidence_interval()
```

```{r KMEP, echo = FALSE, fig.cap="Kaplan Meier plot for EP" }
library(ggsurvfit)
survfit2(Surv ~ EPchar, data = dat) %>% 
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall survival probability"
  ) + 
  add_confidence_interval()
```

Figures \ref{fig:KMEF} and \ref{fig:KMEP} show Kaplan-Meier plots for EF and EP.  Whiel there is a substantial amount of overlap between the K-M curves between the enforcer and non-enforcers groups when enforcers is defined using the EF criteria, the enforcer group when defined by the EP criteria shows lower survival probabilities starting around age 50 relative to the non-enforcer group.  Though it is important to remember that these two groups differ substantially in age, height, and weight.

# Methods {#sec:methods}
Rather than focusing on mortality rates, as in @popkin2023, instead the focus here is to perform a re-analysis of the original study using survival analysis techniques, specifically Cox proportional hazards models, to look for differences in survival times of enforcers versus non-enforcers.  [[Why do we think survival analysis is the correct choice for analyzing this data.  Explain.  ]]

@@

# Results {#sec:results}
Two Cox proportional hazards models @CoxPH were fit: one for each definition of enforcer defined in @Popkin2023 (i.e. E-F and E-P) with time until death as the response variable.  In addition, to indicators for each definition of enforcer, games played (GP), height (Ht), weight (Wt), year of birth (DOB), and position (Center (C), Wing (W), or Defender(D)) were added to each model as control variables.

@@

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon Feb 26 09:06:03 2024 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & $\beta$ & $e^{\beta}$ & s.e. & p-value \\ 
  \hline
E-F TRUE & 0.17328 & 1.18920 & 0.28884 & 0.54856 \\ 
  GP & -0.00059 & 0.99941 & 0.00022 & 0.00815 \\ 
  Ht & -0.04484 & 0.95615 & 0.04430 & 0.31144 \\ 
  Wt & 0.02274 & 1.02300 & 0.00652 & 0.00049 \\ 
  DOB & -0.01766 & 0.98250 & 0.00821 & 0.03140 \\ 
  $I_D$ & -0.36479 & 0.69434 & 0.17428 & 0.03633 \\ 
  $I_W$ & -0.25905 & 0.77178 & 0.15869 & 0.10258 \\ 
   \hline
\end{tabular}
\caption{E-F \label{tab:EF}}
\end{table}

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon Feb 26 09:04:35 2024 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & $\beta$ & $e^{\beta}$ & s.e. & p-value \\ 
  \hline
E-P TRUE & 0.83023 & 2.29385 & 0.26443 & 0.00169 \\ 
  GP & -0.00051 & 0.99949 & 0.00021 & 0.01759 \\ 
  Ht & -0.03859 & 0.96215 & 0.04406 & 0.38111 \\ 
  Wt & 0.02001 & 1.02022 & 0.00654 & 0.00221 \\ 
  DOB & -0.01778 & 0.98238 & 0.00827 & 0.03150 \\ 
  $I_D$ & -0.36900 & 0.69142 & 0.17444 & 0.03440 \\ 
  $I_W$ & -0.28804 & 0.74973 & 0.15946 & 0.07087 \\ 
   \hline
   \hline
\end{tabular}
\caption{E-P \label{tab:EP}}
\end{table}
